---
title: "Class17"
author: "Aileen Andrade (PID A17033749"
date: "03-04-25"
format: pdf
---
Instance ID

i-0b7b517a68441b279 (bimm143_aileenandrade)
Open an SSH client.

Locate your private key file. The key used to launch this instance is bimm143_aileenandrade.pem

Run this command, if necessary, to ensure your key is not publicly viewable.
chmod 400 "bimm143_aileenandrade.pem"

Connect to your instance using its Public DNS:
ec2-54-69-133-115.us-west-2.compute.amazonaws.com

Example:

ssh -i keyfile ubuntu@ec2-54-69-133-115.us-west-2.compute.amazonaws.com

scp -i keyfile ubuntu@ec2-54-69-133-115.us-west-2.compute.amazonaws.com:~/*_quant .

> Q1. File paths correct for data input.

```{r}
library(tximport)
library(rhdf5)

# setup the folder and filenames to read
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

```{r}
head(txi.kallisto$counts)
```

```{r}
colSums(txi.kallisto$counts)
```

```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```

```{r}
pca <- prcomp(t(x), scale=TRUE)
```

```{r}
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```

> Q2. Use ggplot to make a similar figure of PC1 vs PC2 and a seperate figure PC1 vs PC3 and PC2 vs PC3.

```{r}
library(ggplot2)
library(ggrepel)

# Make metadata object for the samples
colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)

# Make the data.frame for ggplot 
y <- as.data.frame(pca$x)
y$Condition <- as.factor(colData$condition)

ggplot(y) +
  aes(PC1, PC2, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

```{r}
# Plot PC1 vs PC2
p1 <- ggplot(y) +
  aes(x = PC1, y = PC2, col = Condition) +
  geom_point(size = 3) +
  geom_text_repel(label = rownames(y)) +
  labs(title = "PCA Plot: PC1 vs PC2") +
  theme_bw()
print(p1)
```

```{r}
# Plot PC1 vs PC3
p2 <- ggplot(y) +
  aes(x = PC1, y = PC3, col = Condition) +
  geom_point(size = 3) +
  geom_text_repel(label = rownames(y)) +
  labs(title = "PCA Plot: PC1 vs PC3") +
  theme_bw()
print(p2)
```

```{r}
# Plot PC2 vs PC3
p3 <- ggplot(y) +
  aes(x = PC2, y = PC3, col = Condition) +
  geom_point(size = 3) +
  geom_text_repel(label = rownames(y)) +
  labs(title = "PCA Plot: PC2 vs PC3") +
  theme_bw()
print(p3)
```

